% !TeX program = pdflatex
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{setspace}
\onehalfspacing

\usepackage{amsmath,amssymb,mathtools,bm}
\usepackage{amsthm}
\numberwithin{equation}{section}
\usepackage{graphicx}
\usepackage{array}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage[round,authoryear]{natbib}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Normal}{\mathcal{N}}
\newcommand{\Exp}{\mathrm{Exp}}
\newcommand{\GIG}{\mathrm{GIG}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\vect}[1]{\bm{#1}}
\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\RHS}{\mathrm{RHS}}
\newcommand{\TableStyle}{\footnotesize\setlength{\tabcolsep}{4.5pt}\renewcommand{\arraystretch}{1.12}}
\DeclareMathOperator{\argmin}{arg\,min}
\newtheorem{proposition}{Proposition}

\title{\vspace{-0.6cm}Bayesian Relaxed Quantile Regression for Direct Interval Forecasting}
\author{Antonio De Leon$^{1}$ \and Raquel Prado$^{1}$ \and Bruno Sans\'o$^{1}$}
\date{\small $^{1}$Department of Statistics, University of California, Santa Cruz\\
\vspace{0.25em}\today}

\begin{document}
\maketitle
\vspace{-1.25em}

\begin{abstract}
\noindent
Prediction intervals are often built from two fitted conditional quantiles, but
that construction fixes endpoint probabilities before assessing how interval
coverage, width, and asymmetry interact. Relaxed quantile regression (RQR)
instead estimates two interval roots directly through a residual-product loss.
We develop a Bayesian generalized-posterior formulation for this target,
described as coverage-targeted interval-root regression under the RQR loss, and
show that a pseudo-residual-normalized exponentiated loss admits a
pseudo-asymmetric-Laplace normal--exponential augmentation. Conditional on one
root and the latent scales, the other root has a Gaussian full conditional.
This yields a partially collapsed Gibbs sampler with either a fixed learning
rate or a learned inverse loss scale. The construction supports fixed nonlinear
readouts, including RQR-DESN, and a linear dynamic extension in which each root
trajectory is updated by forward-filtering backward-sampling. Fixed evolution
covariances, frozen component-discount templates, and shared component-specific
evolution scales define coherent state priors. We show that the analogous
adaptive conditional-discount recursion is generally incompatible with the
advertised pair of Gaussian full conditionals. Posterior summaries are
interval-root functions under a loss-based update, not posterior predictive
response draws.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

Interval forecasts are central when decisions depend on uncertainty ranges
rather than on a conditional mean or a single quantile. A common construction
fits two quantile models, for example at probabilities \(\alpha/2\) and
\(1-\alpha/2\), and reports the resulting lower and upper curves. That
workflow is natural when equal-tailed quantiles are the target, but it fixes
the endpoint probabilities before considering whether an asymmetric interval
could achieve the same coverage with a better score or more appropriate
balance for the application.

Relaxed quantile regression (RQR) provides a different route by estimating two
interval roots directly \citep{PouplinEtAl2024RQR}. For a target coverage
level \(c\in(0,1)\), RQR applies a check loss to the product
\((y-\eta_1)(y-\eta_2)\). The sign of this product records whether the
response lies between the two roots, so the loss targets an interval rather
than a preassigned pair of endpoint quantiles. The original RQR construction
is optimization-based. This paper develops a Bayesian generalized-posterior
route for the same core loss.

The contribution is computational as well as statistical. By exponentiating
the RQR loss in the generalized-Bayes sense of
\citet{BissiriHolmesWalker2016GeneralBayes}, the residual-product loss can be
written with the normal--exponential augmentation used for asymmetric Laplace
quantile regression \citep{yu2001bayesian,KozumiKobayashi2011Gibbs}. The
augmentation is applied to a pseudo-residual, not to the original response.
Conditional on the latent scales and one root, the other root enters linearly.
This gives Gaussian block updates and makes Gibbs sampling feasible for
high-dimensional readouts with ridge or regularized-horseshoe shrinkage
\citep{CarvalhoPolsonScott2010HS,PiironenVehtari2017RHS}.

The fixed-feature version embeds this sampler in nonlinear dynamic
representations. In RQR-DESN, a deep echo state network supplies a deterministic
feature map and posterior computation is performed only on the two interval
readouts \citep{Jaeger2001EchoState,LukoseviciusJaeger2009ReservoirComputing,
GallicchioMicheliPedrelli2018DeepESNDesign,McDermottWikle2018DeepESNUQ}.
The same algebra also suggests a linear dynamic RQR model: conditional on one
root trajectory and the latent scales, the other root trajectory has a Gaussian
state-space observation equation and can be sampled by FFBS. This extension is
important because it places RQR beside dynamic quantile and state-space
forecasting methods while preserving the direct-interval target
\citep{GoncalvesMigonBastos2020DQLM,SinghRaman2025ESNStateSpace}.

The interpretation is deliberately limited. The generalized posterior updates
beliefs about interval endpoint functions under the declared RQR loss and
prior. It is not an ordinary response likelihood and does not define posterior
predictive response draws. Empirical validation should therefore focus on
held-out RQR loss, empirical coverage, width, endpoint recovery when oracle
RQR roots are known, root-switching diagnostics, and sensitivity to the
generalized-Bayes learning rate. The usual interval score is informative as a
secondary comparator but targets equal-tailed intervals and is not neutral
between all intervals having the same coverage
\citep{GneitingRaftery2007,Gneiting2011,GneitingKatzfuss2014,
SyringMartin2019GeneralPosteriorCalibration,Shaby2014OpenFacedSandwich}.

\section{Relaxed Quantile Loss}
\label{sec:rqr-loss}

Let \(y_t\) be the response and let \(\vect x_t\in\R^p\) be a fixed design row.
For a coverage level \(c\in(0,1)\), define two exchangeable root readouts
\[
\eta_{1t}=\vect x_t^\top\vect\beta_1,\qquad
\eta_{2t}=\vect x_t^\top\vect\beta_2.
\]
The reported interval endpoints are ordered after fitting:
\[
L_t=\min(\eta_{1t},\eta_{2t}),\qquad
U_t=\max(\eta_{1t},\eta_{2t}).
\]
The root labels are not lower and upper labels; they are exchangeable
computational labels.

Define the residual product
\[
e_t=(y_t-\eta_{1t})(y_t-\eta_{2t})
\]
and the check loss
\[
\rho_c(e)=e\{c-\ind(e<0)\}.
\]
Then \(e_t<0\) if and only if \(y_t\) lies between the two roots. The empirical
RQR loss is
\[
\mathcal L_{T,c}(\vect\beta_1,\vect\beta_2)
=\sum_{t=1}^T \rho_c(e_t).
\]
\begin{proposition}[Population characterization]
\label{prop:population-characterization}
Fix \(x\) and let \(a<b\). Suppose
\(\E(Y^2\mid X=x)<\infty\), the conditional distribution has no mass at \(a\)
or \(b\), differentiation may be interchanged with conditional expectation,
and \((a,b)\) is an interior stationary point over an unrestricted pointwise
endpoint class. Then the two population first-order conditions imply
\[
\Pr(a<Y<b\mid X=x)=c,
\]
and
\[
\E\{Y\ind(a<Y<b)\mid X=x\}
=c\,\E(Y\mid X=x).
\]
The second equality is equivalently zero conditional covariance between \(Y\)
and the interval-membership indicator.
\end{proposition}

The population coverage implication is due to
\citet{PouplinEtAl2024RQR}. The conditional first-moment equation is a further
consequence of the same first-order system and, to our knowledge, was not
stated in the original article. If the conditional law has endpoint atoms,
the corresponding directional conditions give
\[
\Pr(a<Y<b\mid X=x)\le c\le\Pr(a\le Y\le b\mid X=x),
\]
so exact open-interval coverage requires endpoint continuity. Coincident roots
reduce the loss to \(c(Y-a)^2\) and target the conditional mean rather than a
nontrivial interval.

This motivates RQR as a direct interval target that can adapt to asymmetric
conditional error structure without assigning endpoint quantile levels in
advance. These equations characterize an unrestricted population target only
under the stated conditions. With linear or otherwise restricted root classes,
optimization instead enforces design-weighted score equations; it need not
give conditional coverage at every covariate value. The supplement records
both forms and does not rely on a data-dependent-interval proof.
In particular, we do not use the original article's finite-sample unbiasedness
argument: fitted endpoints are data dependent, and the empirical objective is
nonsmooth when an observation coincides with a root. A valid finite-sample or
asymptotic generalization result requires a separate empirical-process
analysis.

\section{Generalized Posterior and Augmentation}
\label{sec:posterior}

For priors \(\pi_\beta(\vect\beta_1)\) and
\(\pi_\beta(\vect\beta_2)\), define the generalized posterior
\[
\Pi_{\mathrm R}
(d\vect\beta_1,d\vect\beta_2\mid\vect y,\mat X)
\propto
\pi_\beta(d\vect\beta_1)\pi_\beta(d\vect\beta_2)
\exp\{-\omega_{\mathrm R}\mathcal L_{T,c}(\vect\beta_1,\vect\beta_2)\},
\]
where \(\omega_{\mathrm R}>0\) is a learning rate. A learned-scale version
uses a positive reference loss scale \(s_L\), sets
\(\omega_{\mathrm R}=\lambda/s_L\), and places a Gamma prior on \(\lambda\).
We call

\[
\Pi_{\mathrm R}(d\vartheta,d\lambda\mid\vect y)
\propto
\pi(d\vartheta)\lambda^{T}\pi_\lambda(d\lambda)
\exp\{-\lambda\mathcal L_{T,c}(\vartheta)/s_L\},
\]
the \emph{pseudo-residual-normalized learned-scale target}. The factor
\(\lambda^T\) arises from normalizing an asymmetric-Laplace kernel in the
abstract pseudo-residual \(e_t\); it does not normalize a response density in
\(y_t\). Here \(\vartheta\) collects both roots. If
\(\lambda\sim\mathrm{Gamma}(a_\lambda,b_\lambda)\) in shape--rate form, its
collapsed conditional is

\[
\lambda\mid\vartheta,\vect y
\sim\mathrm{Gamma}\left(a_\lambda+T,
b_\lambda+\mathcal L_{T,c}(\vartheta)/s_L\right).
\]
The factor \(\lambda^T\) is part of the declared target rather than an
irrelevant constant when \(\lambda\) is learned. A pure-loss sensitivity target
that omits this factor is kept separate.

The learned \(\lambda\) is part of this declared hierarchical generalized
target. Coherence for a fixed loss scale does not establish that \(\lambda\) is
an ordinary jointly learnable response parameter or that its generalized
posterior calibrates endpoint uncertainty
\citep{LeeLiuNicholls2025LearningRate,WuMartin2020LearningRateComparison}.
We therefore use fixed-rate sensitivity analyses as the primary specification
and treat learned-scale inference as an additional convention. When \(s_L\) is
computed from training responses, its calculation sample and variance
convention are recorded and \(s_L\) is frozen before simulation; the resulting
target is data dependent in an empirical-Bayes sense.

Let
\[
r_t=y_t^2,\qquad
m_t^{\mathrm R}=y_t(\eta_{1t}+\eta_{2t})-\eta_{1t}\eta_{2t}.
\]
Then \(r_t-m_t^{\mathrm R}=e_t\). With
\[
\sigma_{\mathrm R}=s_L/\lambda=\omega_{\mathrm R}^{-1},\qquad
\xi_c=\frac{1-2c}{c(1-c)},\qquad
\phi_c=\frac{2}{c(1-c)},
\]
the exponentiated loss contribution is proportional to
\[
\int_0^\infty
\Normal\!\left(
r_t\mid m_t^{\mathrm R}+\xi_c v_t,\,
\phi_c\sigma_{\mathrm R}v_t
\right)
\Exp(v_t\mid \text{rate}=1/\sigma_{\mathrm R})\,dv_t .
\]
This identity is a computational representation of the loss kernel. It should
not be read as a response model for \(y_t\).

\section{Gibbs Sampling}
\label{sec:gibbs}

Conditional on \(\vect\beta_2\), the pseudo-observation equation is linear in
\(\vect\beta_1\):
\[
\mat A_1=\diag(\vect y-\vect\eta_2)\mat X,\qquad
\vect z_1=\vect y\circ(\vect y-\vect\eta_2)-\xi_c\vect v.
\]
With \(W_v=\diag\{(\phi_c\sigma_{\mathrm R}v_t)^{-1}\}\) and Gaussian prior
canonical parameters \((\mat Q_{01},\vect h_{01})\),
\[
\mat Q_1=\mat Q_{01}+\mat A_1^\top W_v\mat A_1,\qquad
\vect h_1=\vect h_{01}+\mat A_1^\top W_v\vect z_1,
\]
so
\[
\vect\beta_1\mid\cdot\sim
\Normal(\mat Q_1^{-1}\vect h_1,\mat Q_1^{-1}).
\]
The update for \(\vect\beta_2\) is identical after swapping root labels. The
latent scales have generalized inverse Gaussian full conditionals. For the
learned-scale target, one valid partially collapsed scan draws \(\lambda\) from
its collapsed Gamma conditional, refreshes all latent scales using that new
draw, and then updates roots 1 and 2. The refresh order is part of the sampler
contract. Adaptive shrinkage priors use separate shrinkage states for the two
root coefficient vectors.

\section{Model Embeddings}
\label{sec:embeddings}

The Gibbs update applies to any fixed design. In a linear regression embedding,
\(\mat X\) contains ordinary covariates. In RQR-DESN, \(\mat X\) contains
deterministic deep echo state features computed from a recorded reservoir
architecture, seed, scaling rule, and washout convention. The posterior
conditions on those features and updates only the two readout surfaces.

For an RQR-DLM or state-space embedding, each root can be represented by a
latent state path, for example
\[
\eta_{kt}=F_t^\top\theta_{kt},\qquad
\theta_{kt}=G_t\theta_{k,t-1}+\epsilon_{kt},
\qquad \epsilon_{kt}\sim\Normal(0,W_{kt}),
\qquad k=1,2.
\]
The two trajectories can be written as one stacked state with block-diagonal
evolution matrices. That representation does not, however, give one Gaussian
FFBS update. Jointly in the two root states, the augmented observation kernel
contains the square of
\[
(y_t-F_t^\top\theta_{1t})(y_t-F_t^\top\theta_{2t}),
\]
and therefore includes fourth-order cross-terms. Conditional on either complete
root path, the same expression is linear in the other path. The sampler
accordingly partitions the stacked state by root and uses two sequential FFBS
blocks. These are exact full-conditional updates for the declared fixed-joint
target; the issue is potential Markov-chain mixing, not posterior
approximation.

Conditional on the second root path and the RQR latent scales, the first root
path enters through
\[
z_{1t}=y_t(y_t-\eta_{2t})-\xi_c v_t,\qquad
H_{1t}=(y_t-\eta_{2t})F_t^\top,qquad
V_t=\phi_c s_Lv_t/\lambda.
\]
This is a scalar-observation Gaussian state-space system, so the complete root
path can be sampled by forward-filtering backward-sampling. The second root
path is analogous and conditions on the newly drawn first path.

The evolution specification matters to the target. Fixed \(W_{kt}\) defines a
Gaussian state prior and gives exact alternating FFBS. A frozen discount
template first maps component-specific discounts to a complete \(W_{1:T}\)
sequence through a declared reference covariance recursion and then holds that
sequence fixed throughout MCMC. Conditional on that template, this gives an
exact sampler for the stated prior; a template estimated from training data is
recorded as an empirical-Bayes specification. By contrast, recomputing \(W_t\)
from each conditional filter covariance
makes it depend on the other root and latent-scale blocks. We retain that
exdqlm-compatible recursion only as an experimental working update. A
two-time scalar mixed-derivative argument in the supplement shows that its two
advertised conditional FFBS kernels are generally incompatible with a common
positive smooth joint density.

For an exact component-specific alternative, partition the state into blocks
and set
\[
W_t(\vect q)=\operatorname{blockdiag}
\{q_1Q_{1t},\ldots,q_JQ_{Jt}\},
\qquad q_j\sim\operatorname{IG}(a_j,b_j),
\]
where \(Q_{jt}\) are fixed positive-definite templates. The multipliers are
shared across the two roots to preserve prior exchangeability. Sampling the
time-zero states restores the innovation factorization and yields conjugate
inverse-Gamma updates for each \(q_j\); the complete derivation appears in the
supplement.

The implementation distinguishes mathematical target status, numerical
execution, and reproducibility eligibility. Fixed and component-scale modes
have a declared joint target, but they are target-and-numerical eligible only
when every Gaussian factorization succeeds without repair. The default policy
rejects jitter and negative-eigenvalue projection. An audit policy may instead
record the matrix role, time, root and iteration where applicable, jitter, and
eigenvalue diagnostics for every repair. Symmetry and definiteness tolerances
are relative to the actual matrix scale; an exactly zero matrix uses a
separately recorded absolute jitter fallback only under the audit policy.
Promotion additionally requires a clean checkout at an explicitly expected
repository commit.

\section{Validation Plan}
\label{sec:validation}

The standalone simulation study will compare fixed-design RQR, RQR-DESN,
RQR-DLM, quantile-derived intervals, and empirical interval baselines under
matched data-generating mechanisms, samples, train/test splits, seeds, and
forecast windows. Primary summaries are held-out RQR loss, empirical coverage
and coverage error, width at comparable coverage, and endpoint recovery against
oracle RQR roots. Midpoint error, root switching, learned-scale diagnostics,
and the standard interval score are secondary summaries. Results will be
stratified by data-generating mechanism and forecast horizon rather than
collapsed into unsupported global claims. VB may be useful for screening or
initialization, but calibrated VB uncertainty will not be promoted until it
receives its own derivation and validation.

The standard interval score remains secondary because its canonical endpoints
are the corresponding equal-tailed quantiles. RQR does not preassign those
endpoint probabilities. Promotion-grade run manifests require an exact expected
repository commit. Fit objects record the detected commit and clean or unknown
status separately, together with package provenance, data and matrix digests,
numerical policy, RNG state, and a complete continuation checkpoint under a
versioned schema. Continuation verifies schema, data and model digests, and the
recorded numerical environment before claiming bitwise equivalence. Full state
paths are opt-in. Repair counts and numerical eligibility are inherited across
all continuation segments. Each generation records its parent checkpoint,
segment and cumulative repair counts, exactness, reproducibility, promotion
status, mismatch ledger, and requested and resolved FFBS backends; validation
reconstructs these recursions before continuation. Promotion-grade primary and
external namespaces are built in isolated libraries from Git archives whose
file modes, blob identifiers, and paths match the declared commit trees.
Promotion also requires the executing primary namespace to match that isolated
runtime; a clean source checkout alone is insufficient. The local-build
attestation rehashes the build and installation command receipts and logs,
checks the built source-package contents against the verified archive, and
binds the installed runtime to that source package.
Terminal states and component-scale draws are retained for root forecasting
with explicit future covariance templates.

A bounded intercept-only learned-scale fixture cross-checks the production
partially collapsed scan against a separately coded fully augmented scan and
event-boundary-aware root quadrature after analytically integrating the loss
scale. Four
chains per scan satisfy the predeclared rank-normalized convergence and
effective-sample-size gates, all mean and CDF comparisons satisfy their
Monte Carlo error bounds, and no numerical repair is used. This verifies the
implemented small-fixture target; it does not establish frequentist coverage
calibration or authorize the production simulation.

\section{Discussion}
\label{sec:discussion}

RQR-GIBBS is best viewed as a generalized-Bayes workflow for direct interval
targets. Its value is not that it replaces quantile forecasting in all
settings, but that it gives a declared loss-based update over interval roots
when the scientific target is an interval rather than a pair of fixed quantile
levels. The native fixed-design and state-space paths now share the same loss,
scale, and root-ordering contracts. Adaptive discounting has been resolved as a
working rather than exact conditional recursion, while the shared
component-scale construction provides a coherent alternative. Production
evidence remains gated on a frozen matched simulation protocol and explicit
run approval; no response-predictive interpretation follows from the loss
update.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
