% !TeX program = pdflatex
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{setspace}
\onehalfspacing
\usepackage{amsmath,amssymb,mathtools,bm,booktabs,tabularx,array,amsthm}
\usepackage{enumitem}
\usepackage[round,authoryear]{natbib}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Normal}{\mathcal{N}}
\newcommand{\Exp}{\mathrm{Exp}}
\newcommand{\GIG}{\mathrm{GIG}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\vect}[1]{\bm{#1}}
\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\RHS}{\mathrm{RHS}}
\newcommand{\TableStyle}{\footnotesize\setlength{\tabcolsep}{4pt}\renewcommand{\arraystretch}{1.08}}
\newtheorem{proposition}{Proposition}

\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thesubsection}{S\arabic{section}.\arabic{subsection}}

\title{Supplementary Material for \emph{Bayesian Relaxed Quantile Regression for Direct Interval Forecasting}}
\author{Antonio De Leon \and Raquel Prado \and Bruno Sans\'o}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This supplement records the RQR population target, pseudo-AL augmentation,
partially collapsed learned-scale sampler, fixed-design Gaussian blocks,
dynamic FFBS equations, component-discount contracts, and reproducibility
gates for the standalone RQR-GIBBS article.
\end{abstract}

\section{Scope}

The main article presents the statistical target and the core Gibbs sampler.
This supplement keeps the derivations and implementation contracts recoverable.
The RQR update is loss based and does not define an ordinary response
likelihood. Root-trajectory draws from a state evolution equation remain draws
of endpoint functions; they are not posterior predictive draws of the observed
response.

\section{Population Target and Restricted Score Equations}

Fix a covariate value \(x\), write \(a=L(x)<U(x)=b\), and define
\(I_{a,b}(Y)=\ind(a<Y<b)\). Assume
\(\E(Y^2\mid X=x)<\infty\), the conditional risk is differentiable at an
interior stationary point, differentiation and conditional expectation may be
interchanged, and the conditional distribution places no mass at \(a\) or
\(b\). Since
\[
\frac{\partial}{\partial a}(Y-a)(Y-b)=b-Y,
\qquad
\frac{\partial}{\partial b}(Y-a)(Y-b)=a-Y,
\]
the pointwise first-order conditions are
\begin{align}
\E\left[\{c-I_{a,b}(Y)\}(b-Y)\mid X=x\right]&=0,
\label{eq:pop-foc-a}\\
\E\left[\{c-I_{a,b}(Y)\}(a-Y)\mid X=x\right]&=0.
\label{eq:pop-foc-b}
\end{align}
Subtracting \eqref{eq:pop-foc-b} from \eqref{eq:pop-foc-a} and using
\(b-a>0\) gives
\[
\Pr\{a<Y<b\mid X=x\}=c.
\]
Substitution into either first-order condition then gives
\[
\E\{Y\ind(a<Y<b)\mid X=x\}=c\E(Y\mid X=x).
\]

The second equation is equivalently
\(\operatorname{Cov}\{Y,I_{a,b}(Y)\mid X=x\}=0\). It does not imply
independence, symmetry, equal-tailed endpoints, or uniqueness. The coverage
implication is established by \citet{PouplinEtAl2024RQR}; the conditional
first-moment implication is a further consequence of their first-order system.

Endpoint atoms require directional conditions. Write
\(m=(a+b)/2\), \(h=(b-a)/2>0\), so
\(e=(Y-m)^2-h^2\). The left and right derivatives in \(h\) are
\[
D_-R_x(m,h)=2h\{\Pr(a<Y<b\mid x)-c\},
\]
and
\[
D_+R_x(m,h)=2h\{\Pr(a\le Y\le b\mid x)-c\}.
\]
At a local minimum, \(D_-R_x\le0\le D_+R_x\), and therefore
\[
\Pr(a<Y<b\mid X=x)\le c\le\Pr(a\le Y\le b\mid X=x).
\]
If \(a=b=m\), then \(e=(Y-m)^2\) and the risk is
\(c\E\{(Y-m)^2\mid X=x\}\); the coincident-root target is the conditional
mean when the second moment exists.

These implications are conditional, pointwise, and assumption dependent. They
do not follow by treating estimated data-dependent endpoints as fixed. For
linear roots \(\eta_k(X)=X^\top\beta_k\), the corresponding population score
conditions are instead
\begin{align}
\E\left[X\{c-I_{\eta_1,\eta_2}(Y)\}(\eta_2-Y)\right]&=0,\\
\E\left[X\{c-I_{\eta_1,\eta_2}(Y)\}(\eta_1-Y)\right]&=0,
\end{align}
where \(I_{\eta_1,\eta_2}(Y)\) indicates that \(Y\) lies between the
unordered roots. These design-weighted projection equations need not imply
conditional coverage for every \(x\).

We do not use the finite-sample unbiasedness argument in
\citet{PouplinEtAl2024RQR}. Its Monte Carlo step treats fitted,
data-dependent endpoints as fixed, and its zero-gradient argument does not
cover nonsmooth optima at sample points coinciding with a root. A valid
finite-sample or asymptotic result requires a separate empirical-process
analysis.

\section{Pseudo-AL Kernel}

For \(e_t=(y_t-\eta_{1t})(y_t-\eta_{2t})\) and
\(\rho_c(e)=e\{c-\ind(e<0)\}\), a fixed-rate generalized posterior is
proportional to
\(\exp\{-\omega_{\mathrm R}\sum_t\rho_c(e_t)\}\) times the prior. Setting
\[
r_t=y_t^2,\qquad
m_t^{\mathrm R}=y_t(\eta_{1t}+\eta_{2t})-\eta_{1t}\eta_{2t}
\]
gives \(r_t-m_t^{\mathrm R}=e_t\). With
\[
\sigma_{\mathrm R}=\omega_{\mathrm R}^{-1},\qquad
\xi_c=\frac{1-2c}{c(1-c)},\qquad
\phi_c=\frac{2}{c(1-c)},
\]
the normalized asymmetric-Laplace kernel in the pseudo-residual is
\begin{align}
&\frac{c(1-c)}{\sigma_{\mathrm R}}
\exp\{-\rho_c(e_t)/\sigma_{\mathrm R}\}\notag\\
&\quad=\int_0^\infty
\Normal\!\left(e_t\mid\xi_cv_t,\phi_c\sigma_{\mathrm R}v_t\right)
\Exp\!\left(v_t\mid\mathrm{rate}=1/\sigma_{\mathrm R}\right)\,dv_t.
\label{eq:pseudo-al-mixture}
\end{align}
Equation \eqref{eq:pseudo-al-mixture} is an identity for an exponentiated loss
kernel. Its left-hand side is evaluated at a residual product containing
\(y_t\); it is not a sampling density for \(y_t\).

The latent-scale update uses the \(\GIG(\nu,a,b)\) convention proportional to
\(v^{\nu-1}\exp\{-(av+b/v)/2\}\):
\[
v_t\mid\cdot\sim
\GIG\left\{
\frac12,\
\frac{1}{2\sigma_{\mathrm R}c(1-c)},\
\frac{c(1-c)e_t^2}{2\sigma_{\mathrm R}}
\right\}.
\]
At \(e_t=0\), this reduces to
\(v_t\sim\mathrm{Gamma}(1/2,\mathrm{rate}=a/2)\). For \(e_t\ne0\), the native
implementation uses the reciprocal identity
\[
v_t^{-1}\sim\mathrm{IG}\left(\sqrt{a/b_t},a\right),
\]
where the inverse-Gaussian parameters are mean and shape.

\section{Fixed-Design Gaussian Blocks}

Let \(W_v=\diag\{(\phi_c\sigma_{\mathrm R}v_t)^{-1}\}\). Conditional on the
second root, define
\[
\mat A_1=\diag(\vect y-\vect\eta_2)\mat X,\qquad
\vect z_1=\vect y\circ(\vect y-\vect\eta_2)-\xi_c\vect v.
\]
For Gaussian prior canonical parameters \((\mat Q_{01},\vect h_{01})\),
\[
\mat Q_1=\mat Q_{01}+\mat A_1^\top W_v\mat A_1,\qquad
\vect h_1=\vect h_{01}+\mat A_1^\top W_v\vect z_1.
\]
The conditional draw is
\[
\vect\beta_1\mid\cdot\sim
\Normal(\mat Q_1^{-1}\vect h_1,\mat Q_1^{-1}).
\]
The second root uses the same update after exchanging labels. Ridge and
\(\RHS\) priors enter through the same Gaussian canonical form, with separate
shrinkage states for each root. The current native dynamic implementation uses
Gaussian state priors; fixed-design RHS-family support remains an explicit
adapter to the pinned exdqlm implementation until it is promoted natively.

\section{Learned Inverse-Loss Scale}

Let \(s_L>0\) be a declared reference loss scale and set
\(\sigma_{\mathrm R}=s_L/\lambda\). The normalized target is
\[
\pi(\vartheta,\lambda\mid\vect y)
\propto
\pi(\vartheta)\pi_\lambda(\lambda)
\lambda^n\exp\{-\lambda\mathcal L_{n,c}(\vartheta)/s_L\},
\label{eq:normalized-scale-target}
\]
where \(\vartheta\) denotes both roots. For
\(\lambda\sim\mathrm{Gamma}(a_\lambda,b_\lambda)\) in shape--rate form,
\[
\lambda\mid\vartheta,\vect y
\sim\mathrm{Gamma}\left(a_\lambda+n,
b_\lambda+\mathcal L_{n,c}(\vartheta)/s_L\right).
\label{eq:lambda-collapsed}
\]

We call \eqref{eq:normalized-scale-target} the
\emph{pseudo-residual-normalized learned-scale target}. The factor
\(\lambda^n\) normalizes an asymmetric-Laplace kernel in the abstract
pseudo-residual \(e_t\), not a density in \(y_t\). It is an analyst-declared
joint generalized target rather than a consequence that \(\lambda\) is an
ordinary response parameter. Coherence of a fixed-rate loss update does not by
itself establish same-data learning or calibration of this scale
\citep{LeeLiuNicholls2025LearningRate,WuMartin2020LearningRateComparison}.
Fixed-rate sensitivity grids are therefore primary. If \(s_L\) is estimated
from the training responses, the calculation convention is recorded and the
value is frozen before MCMC; this is a data-dependent empirical-Bayes
specification.

If one conditions additionally on \(\vect v\), the normalized augmented
conditional is
\begin{align}
\lambda\mid\vartheta,\vect v,\vect y
\sim\mathrm{Gamma}\Bigg(&a_\lambda+\frac{3n}{2},\notag\\
&b_\lambda+\frac{1}{s_L}\sum_{t=1}^n
\left[v_t+\frac{(e_t-\xi_cv_t)^2}{2\phi_cv_t}\right]\Bigg).
\label{eq:lambda-augmented}
\end{align}

The diagnostic pure-loss target omits \(\lambda^n\). Its collapsed shape is
\(a_\lambda\), and a compatible augmented representation has shape
\(a_\lambda+n/2\). These alternatives are different targets, not two
implementations of the same posterior.

The implemented partially collapsed scan uses the following order:
\begin{enumerate}[leftmargin=2.2em]
\item draw \(\lambda\) from \eqref{eq:lambda-collapsed} using the current roots;
\item refresh every \(v_t\) from its GIG conditional using the new \(\lambda\);
\item update root 1 conditional on root 2 and the refreshed scales;
\item update root 2 conditional on the new root 1 and the refreshed scales.
\end{enumerate}
The second step cannot be skipped after the collapsed draw.

\section{RQR-DESN Contract}

RQR-DESN conditions on a deterministic DESN feature matrix. The reservoir seed,
depth, units, lag window, spectral radius, leak rate, input scaling, washout,
and feature reducers are design variables, not posterior unknowns. Posterior
summaries are ordered endpoint functions, widths, and midpoints.

\section{RQR-DLM Contract and FFBS}

For root \(k=1,2\), let
\[
\eta_{kt}=F_t^\top\theta_{kt},\qquad
\theta_{k0}\sim\Normal(m_0,C_0),\qquad
\theta_{kt}=G_t\theta_{k,t-1}+\epsilon_{kt},\qquad
\epsilon_{kt}\sim\Normal(0,W_{kt}).
\]
Conditional on \(\theta_{2,1:T}\), \(v_{1:T}\), and \(\lambda\), root 1 has
\begin{align}
z_{1t}&=H_{1t}\theta_{1t}+\varepsilon_{1t},\\
z_{1t}&=y_t(y_t-\eta_{2t})-\xi_cv_t,\\
H_{1t}&=(y_t-\eta_{2t})F_t^\top,\\
\varepsilon_{1t}&\sim\Normal(0,V_t),\qquad
V_t=\phi_cs_Lv_t/\lambda.
\end{align}
Root 2 is obtained by exchanging labels.

For either conditional Gaussian system, write \(m_{t-1},C_{t-1}\) for filtered
moments. The forward recursion is
\begin{align}
a_t&=G_tm_{t-1}, & P_t&=G_tC_{t-1}G_t^\top,\\
R_t&=P_t+W_t, & q_t&=H_tR_tH_t^\top+V_t,\\
A_t&=R_tH_t^\top/q_t, & m_t&=a_t+A_t(z_t-H_ta_t),\\
C_t&=R_t-A_tq_tA_t^\top.
\end{align}
Starting from \(\theta_T\sim\Normal(m_T,C_T)\), the backward draw for
\(t=T-1,\ldots,1\) uses
\begin{align}
B_t&=C_tG_{t+1}^\top R_{t+1}^{-1},\\
h_t&=m_t+B_t(\theta_{t+1}-a_{t+1}),\\
H_t^{\mathrm B}&=C_t-B_tR_{t+1}B_t^\top,\\
\theta_t\mid\theta_{t+1},\cdot&\sim\Normal(h_t,H_t^{\mathrm B}).
\end{align}

For fixed evolution covariances, the implementation integrates
\(\theta_{k0}\) out of the root-path update. Thus
\[
a_{k1}=G_1m_0,\qquad
R_{k1}=G_1C_0G_1^\top+W_1,
\]
and backward sampling terminates at \(t=1\). The component-scale extension
samples \(\theta_{k0}\) separately because its first innovation is required by
the conjugate scale update.

Missing responses skip the measurement update. The implementation uses
Cholesky factorization and does not silently turn an indefinite covariance into
a positive one by taking absolute singular values. Under the default fail-fast
policy, a failed factorization stops the fit. The audit policy uses a declared
jitter ladder and records every covariance alteration. When a backward prior
covariance is repaired, the repaired matrix is used consistently in the gain,
smoothing covariance, and path-draw covariance. Fixed-target fits with nonzero
repair counts are not eligible for promotion-grade exact-target analyses.

\section{Component-Specific Discount Factors}

Let component discounts be \(\delta_1,\ldots,\delta_J\), with
\(0<\delta_j\le1\), and let their state dimensions sum to \(p\). Define the
\(p\times p\) matrix \(D\) to have value
\((1-\delta_j)/\delta_j\) throughout the within-component block \(j\) and zero
between components. This reproduces the public \texttt{df}/\texttt{dim.df}
construction in exdqlm 1.1.0. The covariance recursion is
\[
P_t=G_tC_{t-1}G_t^\top,\qquad
W_t=D\circ P_t,\qquad
R_t=P_t+W_t,
\]
where \(\circ\) denotes an elementwise product. Thus a within-component
covariance block is inflated by \(1/\delta_j\), whereas cross-component
covariances are not inflated.

The standalone API distinguishes three uses:
\begin{enumerate}[leftmargin=2.2em]
\item \texttt{fixed\_W}: \(W_{1:T}\) is supplied and fixed. This gives exact
FFBS for the declared Gaussian state prior.
\item \texttt{discount\_template}: a reference design and reference observation
variance generate \(W_{1:T}\) once through the recursion above. The sequence is
then frozen before MCMC. This also gives exact FFBS for the resulting declared
prior, conditional on the frozen template. If the reference quantities are
estimated from the training responses, the resulting prior specification is
empirical Bayes and must be recorded as such.
\item \texttt{adaptive\_discount}: \(W_t\) is recomputed inside each conditional
filter. In RQR, \(H_t\) and \(V_t\) depend on the other root, \(v_t\), and
\(\lambda\), so the implied evolution covariance also depends on those blocks.
The current alternating kernel omits terms that would generally arise from
such dependence in a fixed joint density. This mode is therefore experimental
and is not described as exact Gibbs for the stated generalized posterior.
\end{enumerate}
The same discounts are used for both roots unless a future model deliberately
breaks their prior exchangeability.

\subsection{Incompatibility of adaptive conditional discounts}

The experimental designation can be made precise. Consider \(T=2\), \(p=1\),
\(F_t=G_t=1\), and a nonzero discount inflation \(d=(1-\delta)/\delta\).
For the root-1 conditional filter, the time-1 covariance and time-2 innovation
variance include
\[
C_{11}=\frac{R_{11}V_1}
{(y_1-\theta_{21})^2R_{11}+V_1},
\qquad
w_1(\theta_{21})=dC_{11}.
\]
The conditional log density therefore contains
\[
-\frac12\log w_1(\theta_{21})
-\frac{(\theta_{12}-\theta_{11})^2}{2w_1(\theta_{21})},
\]
whose mixed derivative is generically
\[
\frac{\partial^2}{\partial\theta_{21}\partial\theta_{12}}
\log\pi(\vect\theta_1\mid\vect\theta_2,\cdot)
=
\frac{(\theta_{12}-\theta_{11})w_1'(\theta_{21})}
{w_1(\theta_{21})^2}\ne0.
\]
In the advertised root-2 conditional, the adaptive variance depends on
\(\theta_{11}\), not \(\theta_{12}\); its time-1 pseudo-observation does not
involve \(\theta_{12}\), and its time-2 pseudo-observation does not involve
\(\theta_{21}\). Hence the corresponding mixed derivative is zero. A positive
twice-differentiable joint density would require equality of these derivatives.
Except in degenerate cases such as \(d=0\), the two conditional adaptive FFBS
kernels are therefore incompatible with a common smooth joint density while
retaining their advertised Gaussian forms.

\subsection{Exact shared component scales}

Partition the state into \(J\) blocks of dimensions \(d_j\), let \(S_j\)
select block \(j\), and fix positive-definite templates \(Q_{jt}\). Define
\[
W_t(\vect q)=\operatorname{blockdiag}
\{q_1Q_{1t},\ldots,q_JQ_{Jt}\},
\qquad
q_j\sim\operatorname{IG}(a_j,b_j),
\]
where the inverse-Gamma density is proportional to
\(q_j^{-a_j-1}\exp(-b_j/q_j)\). The same \(q_j\) is shared by the two roots.
For
\[
d_{kjt}=S_j\{\theta_{kt}-G_t\theta_{k,t-1}\},
\]
the full conditional is
\[
q_j\mid\cdot\sim\operatorname{IG}\left(
a_j+Td_j,
b_j+\frac12\sum_{k=1}^2\sum_{t=1}^T
d_{kjt}^\top Q_{jt}^{-1}d_{kjt}
\right).
\]
The \(Td_j\) shape increment includes two roots, each contributing
\(Td_j/2\).

Because the fixed-\(W\) FFBS implementation integrates out the time-zero
state, the component-scale sampler restores it after each root-path draw. Its
conditional moments are
\begin{align}
C_{k0}^\star
&=\{C_0^{-1}+G_1^\top W_1(\vect q)^{-1}G_1\}^{-1},\\
m_{k0}^\star
&=C_{k0}^\star\{C_0^{-1}m_0+
G_1^\top W_1(\vect q)^{-1}\theta_{k1}\},
\end{align}
and
\[
\theta_{k0}\mid\theta_{k1},\vect q
\sim\Normal(m_{k0}^\star,C_{k0}^\star).
\]
One exact sweep updates collapsed \(\lambda\), refreshes \(\vect v\), samples
each root path and its time-zero state, updates all \(q_j\), and optionally
swaps the two complete root labels. Static \(q_j=0\) components are separate
degenerate models and are not part of the inverse-Gamma support.

\section{Validation Gates}

Before a production simulation, the package must pass pure-R/C++ filtering and
smoothing parity against an independent dense Gaussian reference,
multiblock discount construction, missing-observation, extreme-scale GIG,
learned-scale, component-scale conditional, checkpoint-continuation,
root-label, and small end-to-end tests. Promotion-grade exact-target fits must
have zero numerical repairs. A run manifest must then freeze data-generating
mechanisms, sample sizes, train/test windows, forecast horizons, method
configurations, seeds, package versions, and repository commits. Fit objects
record data and matrix digests, RNG state, compiler and BLAS provenance, and a
versioned schema. Heavy state draws and fitted objects remain in ignored local
output directories.

Held-out RQR loss, empirical coverage error, and width at comparable coverage
are primary. Oracle endpoint recovery is primary when the population RQR roots
are available. The standard interval score is secondary because its canonical
target is an equal-tailed interval, whereas RQR does not preassign endpoint
quantile probabilities.

\section{Migrated Preliminary RQR-DESN Table}

The table below is migrated from the Q-DESN supplement as preliminary,
calibration-screened evidence. It is not yet the final standalone simulation
study because it does not include the planned RQR-DLM comparison.

\input{tables/rqr_desn_targeted_confirmation_winner_map.tex}

\nocite{PouplinEtAl2024RQR,BissiriHolmesWalker2016GeneralBayes,KozumiKobayashi2011Gibbs,GoncalvesMigonBastos2020DQLM}
\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
