---
title: "Ordinary RQR v1: fixed, frozen-feature, and dynamic models"
author: "Antonio De Leon, Raquel Prado, and Bruno Sanso"
date: "2026-07-26"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ordinary RQR v1: fixed, frozen-feature, and dynamic models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.5
)
library(rqrgibbs)
```

Ordinary relaxed quantile regression (RQR) updates two interval-root
functions under the residual-product loss. The resulting distribution is a
loss-based generalized posterior. It is not an ordinary likelihood for the
response, and its endpoint draws are not posterior-predictive response draws.

This vignette illustrates three native interfaces:

1. fixed-design regression;
2. a readout conditional on a frozen deep echo-state-network (DESN) feature
   design; and
3. dynamic linear interval-root states.

The deliberately short chains below exercise the interfaces. They are not
sufficient for scientific inference or convergence assessment.

## Fixed-design interval-root regression

The response may contain `NA`; those rows remain in the design but are omitted
from the loss and latent-scale updates.

```{r fixed-data}
set.seed(260726)
n <- 24L
x <- seq(-1, 1, length.out = n)
X <- cbind(intercept = 1, x = x)
y <- 0.4 - 0.6 * x + 0.15 * stats::rt(n, df = 5)
y[c(7, 18)] <- NA_real_
```

A ridge prior is a proper Gaussian prior shared by the two exchangeable raw
roots. Here the generalized-Bayes learning rate is fixed in advance.

```{r fixed-fit}
ridge_prior <- rqr_beta_prior(
  "ridge",
  ridge = list(tau2 = 10)
)

fixed_fit <- rqr_mcmc_fit(
  y = y,
  X = X,
  coverage_level = 0.80,
  learning_rate = 1,
  learning_rate_mode = "fixed_rate",
  beta_prior_obj = ridge_prior,
  numerical_policy = "fail",
  mcmc_control = list(
    n_burn = 40L,
    n_mcmc = 80L,
    thin = 1L,
    seed = 260727L
  )
)
```

`predict_interval()` evaluates ordered endpoint functionals at supplied design
rows. The output flag makes the interpretation explicit.

```{r fixed-summary}
fixed_draws <- rqr_posterior_draws(fixed_fit)
fixed_interval <- predict_interval(fixed_fit, X_new = X)

head(data.frame(
  x = x,
  lower = fixed_interval$lower_mean,
  upper = fixed_interval$upper_mean,
  width = fixed_interval$width_mean
), 4)

stopifnot(
  identical(
    fixed_draws$schema_version,
    "rqrgibbs_static_draws/1.0.0"
  ),
  identical(
    fixed_interval$schema_version,
    "rqrgibbs_interval_prediction/2.0.0"
  ),
  identical(fixed_interval$response_predictive_draws, FALSE)
)
```

Exact same-environment continuation restores the complete checkpoint and RNG
state. Burn-in cannot be added during continuation.

```{r fixed-continuation}
continued_fit <- rqr_mcmc_continue(
  fixed_fit,
  n_mcmc = 20L
)

continued_fit$checkpoint_state$completed_iterations
```

The alternative promoted rate mode is
`"learned_pseudoresidual_normalized"`. Its Gamma update uses the observed-row
count and the declared loss reference scale. The sampled rate is an inverse
loss scale, not a response variance or coverage-calibration parameter.

```{r learned-rate-interface, eval=FALSE}
learned_fit <- rqr_mcmc_fit(
  y = y,
  X = X,
  coverage_level = 0.80,
  learning_rate_mode = "learned_pseudoresidual_normalized",
  loss_reference_scale = 1,
  lambda_prior = list(shape = 4, rate = 4),
  beta_prior_obj = ridge_prior,
  numerical_policy = "fail",
  mcmc_control = list(
    n_burn = 1000L,
    n_mcmc = 3000L,
    seed = 260729L
  )
)
```

Full-Gaussian and native shrunken-shoulder horseshoe priors use the same
fixed-design sampler. The latter requires an explicitly named constant-one
intercept and is available only for ordinary, zero-tilt RQR.
Gaussian coefficient binding is deliberately unambiguous: if either the prior
mean or prior matrix is named, complete identical names are required on both
matrix axes and on a supplied mean, and their order must exactly match
`colnames(X)`. If every Gaussian-prior input is unnamed, binding is explicitly
positional.

```{r prior-contracts}
gaussian_prior <- rqr_beta_prior(
  "gaussian",
  gaussian = list(
    mean = c(0, 0),
    covariance = diag(c(10, 2))
  )
)

rhs_ns_prior <- rqr_beta_prior(
  "rhs_ns",
  rhs_ns = list(
    intercept_name = "intercept",
    tau0 = 0.5,
    zeta2_fixed = 4
  )
)

c(gaussian_prior$type, rhs_ns_prior$type)
```

## Ordinary RQR conditional on frozen DESN features

RQR-DESN uses the same coefficient sampler after a deterministic feature
matrix has been constructed and frozen. The feature-design contract records
its ordered columns, builder, reservoir digest, time alignment, and causal
semantics. It does not run or refit a reservoir during MCMC.

For a self-contained example, the next helper materializes two fixed recurrent
features from a known driver. In an application, an upstream DESN workflow
would supply the corresponding feature matrix and metadata.

```{r frozen-feature-builder}
reservoir_spec <- list(
  layer1_memory = 0.45,
  layer1_input = 0.65,
  layer2_memory = 0.30,
  layer2_input = 0.70
)

materialize_features <- function(driver, initial_state = c(0, 0)) {
  state <- matrix(NA_real_, nrow = length(driver), ncol = 2L)
  previous <- initial_state
  for (tt in seq_along(driver)) {
    layer1 <- tanh(
      reservoir_spec$layer1_memory * previous[1] +
        reservoir_spec$layer1_input * driver[tt]
    )
    layer2 <- tanh(
      reservoir_spec$layer2_memory * previous[2] +
        reservoir_spec$layer2_input * layer1
    )
    state[tt, ] <- c(layer1, layer2)
    previous <- state[tt, ]
  }
  list(
    X = cbind(
      intercept = 1,
      driver = driver,
      layer1 = state[, 1],
      layer2 = state[, 2]
    ),
    terminal_state = previous
  )
}

driver <- sin(seq_len(n + 3L) / 4)
training_features <- materialize_features(driver[seq_len(n)])
reservoir_metadata <- list(
  digest = digest::digest(
    reservoir_spec,
    algo = "sha256",
    serialize = TRUE
  ),
  architecture = "illustrative_two_layer_fixed_reservoir",
  fixed = TRUE
)
```

The frozen interface needs no executing `exdqlm` namespace. A hand-built
design such as this one is deliberately non-promotable. Promotion-grade
reference materialization is a separate design-only route, orchestrated by
`application/scripts/28_materialize_rqr_ordinary_v1_desn_design.R` from exact
isolated `rqrgibbs` and pinned `exdqlm` runtimes.

```{r frozen-design-fit}
desn_design <- rqr_desn_design(
  X = training_features$X,
  y = y,
  time_index = seq_len(n),
  intercept = "intercept",
  builder = list(
    id = "vignette_deterministic_feature_builder",
    version = "1.0.0"
  ),
  reservoir = reservoir_metadata,
  driver = list(
    type = "known_deterministic_driver",
    response_simulation = FALSE
  )
)

desn_fit <- rqr_desn_fit(
  design = desn_design,
  design_engine = "frozen",
  coverage_level = 0.80,
  learning_rate = 1,
  learning_rate_mode = "fixed_rate",
  mcmc_args = list(
    beta_prior_obj = rqr_beta_prior(
      "ridge",
      ridge = list(tau2 = 10)
    ),
    mcmc_control = list(
      n_burn = 40L,
      n_mcmc = 80L,
      thin = 1L,
      seed = 260728L
    )
  )
)

desn_draws <- rqr_posterior_draws(desn_fit)
stopifnot(
  identical(
    desn_fit$schema_version,
    "rqrgibbs_desn_fit/1.2.0"
  ),
  identical(
    desn_draws$schema_version,
    "rqrgibbs_desn_draws/1.0.0"
  ),
  identical(
    desn_fit$model_spec$
      design_materialization_reproducibility_eligible,
    FALSE
  )
)
```

Future features require their own versioned contract. The following rows are
precomputed at the forecast origin by continuing the fixed feature recursion
along the known future driver path.

```{r frozen-future}
future_features <- materialize_features(
  driver[n + seq_len(3L)],
  initial_state = training_features$terminal_state
)

future_design <- rqr_desn_future_design(
  parent_design = desn_design,
  X = future_features$X,
  time_index = n + seq_len(3L),
  semantics = "precomputed_design",
  driver = list(
    origin_fixed = TRUE,
    uses_realized_post_origin_history = FALSE,
    source = "known_deterministic_driver",
    response_simulation = FALSE
  )
)

future_interval <- forecast_paths(
  desn_fit,
  H = 3L,
  future_design = future_design
)

data.frame(
  time = future_design$time_index,
  lower = future_interval$lower_mean,
  upper = future_interval$upper_mean,
  width = future_interval$width_mean
)

stopifnot(
  identical(
    future_interval$schema_version,
    "rqrgibbs_desn_prediction/1.0.0"
  ),
  identical(future_interval$evaluation_semantics, "precomputed_design"),
  identical(future_interval$response_predictive_draws, FALSE)
)
```

The future contract also distinguishes rolling teacher-forced one-step
evaluation from an origin-fixed external-driver path. Those semantics must be
declared when the feature rows are constructed; neither mode creates a
response-simulation distribution.

## Dynamic linear interval roots

RQR-DLM replaces static root coefficients with two exchangeable state paths.
The paths share the same observation design, state evolution, and prior
specification. Although their stacked state prior is Gaussian, the augmented
observation kernel is not jointly quadratic in both root paths. The sampler
therefore alternates two root-specific Gaussian FFBS updates, each conditional
on the current opposing path. These are exact blocked Gibbs updates for a
declared fixed joint target; they are not one simultaneous Gaussian FFBS draw.

The following interface uses an explicit fixed evolution covariance. It is
shown without evaluation because the short chains elsewhere in this vignette
are intended only as interface checks.

```{r dynamic-fixed-w, eval=FALSE}
dynamic_model <- rqr_polytrend(
  order = 2L,
  m0 = c(0, 0),
  C0 = diag(c(4, 1))
)

dynamic_fit <- rqr_dlm_fit(
  y = y,
  model = dynamic_model,
  coverage_level = 0.80,
  evolution_spec = rqr_evolution_fixed(
    diag(c(0.04, 0.01))
  ),
  learning_rate = 1,
  learning_rate_mode = "fixed_rate",
  numerical_policy = "fail",
  mcmc_control = list(
    n_burn = 1000L,
    n_mcmc = 3000L,
    thin = 1L,
    seed = 260730L,
    backend = "cpp"
  )
)

dynamic_interval <- predict_interval(dynamic_fit)

dynamic_draws <- rqr_posterior_draws(dynamic_fit)
stopifnot(
  identical(
    dynamic_fit$provenance$schema_version,
    "rqrgibbs_fit/1.14.0"
  ),
  identical(
    dynamic_draws$schema_version,
    "rqrgibbs_dlm_draws/1.0.0"
  ),
  identical(
    dynamic_interval$schema_version,
    "rqrgibbs_dlm_prediction/1.0.0"
  )
)
```

Three evolution specifications define exact fixed-joint generalized-posterior
samplers when the numerical policy records no repair:

1. `rqr_evolution_fixed()` supplies an explicit fixed covariance;
2. `rqr_freeze_discount_template()` constructs component-specific discount
   covariances before MCMC and then holds that template fixed; and
3. `rqr_evolution_component_scale()` uses fixed component templates with
   shared inverse-Gamma evolution multipliers learned across the two roots.

By contrast, `rqr_evolution_adaptive_working()` recomputes a conditional
discount evolution during the scan. That recursion is explicitly experimental
and does not define the same fixed joint target as the three modes above.

The component-scale constructor and its kernel controls can be checked without
running a chain. Components are stacked into a single state, while their
evolution templates and names retain the same order.

```{r dynamic-component-scale-contract}
component_model <- rqr_polytrend(
  order = 1L,
  C0 = matrix(4, 1, 1),
  name = "level"
) + rqr_regression(
  X = cbind(x = x),
  C0 = matrix(2, 1, 1),
  name = "regression"
)

component_evolution <- rqr_evolution_component_scale(
  templates = list(
    matrix(1, 1, 1),
    matrix(1, 1, 1)
  ),
  component_dims = component_model$component_dims,
  component_names = component_model$component_names,
  prior = list(shape = c(2, 2), rate = c(1, 1))
)

component_kernel_control <- list(
  component_scale_collapsed_update = TRUE,
  component_scale_interweave = TRUE,
  component_scale_slice_width = 1,
  component_scale_slice_max_steps = 100L,
  component_scale_slice_max_shrink = 1000L,
  component_scale_slice_sweeps = 1L,
  component_scale_interweave_cycles = 1L
)

stopifnot(
  identical(component_model$component_names, c("level", "regression")),
  identical(component_evolution$component_names, c("level", "regression"))
)
```

Those seven controls change the transition kernel. Their values, the
component order, the fixed one-root collapsed orientation, time-zero state
completion, scan order, and the final probability-0.5 global root swap are
bound into checkpoints and continuation history. They cannot be changed by
continuing an existing fit. The default component-scale transition uses the
centered inverse-Gamma update; collapsed and interwoven updates are optional
exact alternatives. Every accepted exact DLM mode completes the time-zero
state for both roots whether or not full state paths are stored; storage
therefore does not change the transition or RNG stream. The experimental
adaptive working mode omits time-zero completion and records that omission in
its transition contract.

Future evaluation propagates only the root states. It requires an explicit
future observation design, transition specification, and either fixed future
covariances or fixed future component templates.

```{r dynamic-future-roots, eval=FALSE}
future_roots <- rqr_forecast_roots(
  dynamic_fit,
  FF_future = matrix(c(1, 0), nrow = 2L, ncol = 4L),
  GG_future = matrix(c(1, 1, 0, 1), 2L, 2L, byrow = TRUE),
  W_future = diag(c(0.04, 0.01)),
  seed = 260731L
)

stopifnot(grepl(
  "no response simulation",
  future_roots$interpretation,
  fixed = TRUE
))
stopifnot(identical(
  future_roots$schema_version,
  "rqrgibbs_dlm_forecast/1.0.0"
))
```

The returned trajectories are interval-root state draws conditional on the
declared future evolution. They are not posterior-predictive response draws.
